---
title: "Lab 7 Solved: If statements and conditionals"
author: "Harris Coding Camp"
date: "Summer 2023"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
mrc_data <- haven::read_dta('https://opportunityinsights.org/wp-content/uploads/2018/04/mrc_table10.dta')
```

## I. Warm-up Questions

1. Without running the code, predict what the output will be. Then, see if you were right by running the code in the console.

  **True or False**

  a. `TRUE | FALSE`
  a. `TRUE | (FALSE & TRUE)`
  a. `TRUE | (10 < 4)`
  a. `TRUE | (1 & 0)`
  a. `TRUE | (4 > pi & 3 < pi & exp(1) >= 3 & 1e6 < 2^30)`
  a. `4 > 2 | 2 > 4`
  a. `TRUE | NA`
  a. What rule do these problems demonstrate?
  
  **True and False**

  a. `TRUE & FALSE`
  a. `TRUE & (FALSE & TRUE)`
  a. `TRUE & (10 < 4)`
  a. `TRUE & (1 & 0)`
  a. `TRUE & (4 > pi & 3 < pi & exp(1) >= 3 & 1e6 < 2^30)`
  a. `4 > 2 & 2 > 4`
  a. `NA & FALSE`
  a. What rule do these problems demonstrate?
  
  **True and NA**
  
  a. There are a few times when `NA` are not contagious! Given your analysis above, predict, test and explain the output of the following code.
      
```{r, eval = FALSE}
TRUE & NA    
FALSE & NA    
TRUE | NA    
FALSE | NA    
```

  *Solution*: The big idea here is that sometimes missing information will not change the outcome. For `|`, a single `TRUE` implies the expression is `TRUE`. With `&` a single `FALSE` implies the expression is `FALSE`. In these situations, `NA` can be ignored.

2. Without running the code, predict what the output will be. Then, see if you were right by running the code in the console.

    ```{r, eval = FALSE}
    ifelse(TRUE, "yes", "no")
    ifelse(FALSE, "yes", "no")
    ifelse(c(TRUE, FALSE, TRUE, FALSE), "yes", "no")
    ifelse(c(NA, TRUE, FALSE), "yes", "no")
    ifelse(c(NA, NA, TRUE, FALSE), "yes", "no")
    ifelse(c(NA, NA, TRUE, FALSE) & TRUE, "yes", "no")
    ifelse(c(NA, NA, TRUE, FALSE) | TRUE, "yes", "no")
    ```

3. What do the following do? Predict the output and then run the code.

```{r, eval = FALSE}
# x and y are integer vectors of length 10
x <- c(2:10, NA)
y <- c(NA, 10:2)

# also predict how long will the output be and what data type will we have?
ifelse(x < y, y, x)
ifelse(is.na(x), 0, x)
ifelse(x == 2, paste0(x, "nd"),
       ifelse(x == 3, paste0(x, "rd"),
              paste0(x, "th")))
y  +  ifelse(x > 6, 1, 2)
ifelse(y %% 2 == 0, "even", "odd")
```

4. Write your own `ifelse()` statements using `x` or `y` as input that do the following:

a. Use `&` or `|`.
b. Return a character vector with no `NA`s.
c. Uses nested `ifelse`.
d. All three things at the same time.

There are no correct answers. The idea here is to explore the code to build understanding!


```{r, include = TRUE, eval = FALSE}
# EXample Solutions
ifelse(x > y & !is.na(y), 1, 0)
ifelse(is.na(y) | is.na(x), "missing", "available")
ifelse(y > 7, 
       ifelse(x < 5, TRUE, FALSE),  
       NA) # if y is <= 7 we return NA
ifelse(is.na(y))
```



## Data and background

For this lab, we will use data from Opportunity Insights: https://opportunityinsights.org/data/ 

1. Download the Stata file and Readme for "College Level Characteristics from the IPEDS Database and the College Scorecard".
1. Load it. We will refer to the data as `mrc_data`.

You should be able to read the data with no error in a more-or-less tidy format! (Finally a clean data set!)

1. What does each row represent? How many are there?
1. What columns are present? How many are there? Are the names clear? (The README can help clarify information).


## II. Common uses of `ifelse`

1. Run the following code and you will see the distinct `tier_names` available in the dataset. 

```{r, eval=FALSE}
mrc_data %>% distinct(tier_name)
```

  a. `ifelse` can be used to adjust entries in the `tier_name` column. Change "Two-year (public and private not-for-profit)"
      to "Two-year (public and private)".^[Hint: In the first position, put a condition testing if `tier_name` matches the string. If it does, we replace the string with "Two-year (public and private)", otherwise keep the same data.] 
        
```{r, eval=FALSE}
# SOLUTION
mrc_data %>%
mutate(tier_name = ifelse(tier_name == "Two-year (public and private not-for-profit)",
                           "Two-year (public and private)", 
                           tier_name))
```
    
  b. `ifelse` is often used to collapse tiers. Redefine `tier_name` so that "Nonselective four-year public" and "Nonselective four-year private not-for-profit" are grouped together as "Nonselective four-year (public and private)".^[Hint: The code will be very similar to the previous problem.]

```{r, eval=FALSE}
# SOLUTION
mrc_data %>%
mutate(tier_name = ifelse( tier_name == "Nonselective four-year private not-for-profit"|
                            tier_name == "Nonselective four-year public", 
                           "Nonselective four-year (public and private)", 
                           tier_name))
```    
    
    
2. As you can see below, there are 1466 colleges missing average SAT scores. Sometimes we want to replace `NA`s with a value. For example, linear regressions will drop any row with `NA`s, and we might not want that.

```{r}
mrc_data %>%
  summarise(missing_sat_2013 = sum(is.na(sat_avg_2013)))
```

To avoid dropping rows, sometimes people replace the `NA` with the mean and add a new column that is an indicator of missingness.  Here's a small example of what we expect. 

Fill `NA` in `sat_avg_2013` with the average SAT score of the other colleges and create a column called `missing_sat_avg_2013` that is 1 if `NA` and 0 otherwise.


```{r, eval=TRUE}
# SOLUTION
mrc_data %>%
mutate(missing_sat_avg_2013 = ifelse(is.na(sat_avg_2013), 1, 0),
       sat_avg_2013 = ifelse(is.na(sat_avg_2013), 
                             mean(sat_avg_2013, na.rm = TRUE),
                             sat_avg_2013)) %>%
  select(missing_sat_avg_2013, sat_avg_2013) %>%
  head()
```   

## III. College choice

Imagine the situation: It's 2014 and a group of high school friends want to go to college together. They need to find a college that meets all their preferences. Your job is to find the perfect college.

| Name | SAT Score | Preferences|
| --- | --- | ------ | 
| A-plus Abdul | 1430 | Either ivy plus tier or a flagship school |
| Snooty Sam | 1450 | not a public school |
| Nourishing Nancy | 1590 | school in the midwest so she can be near her grandma |
| Radical Rei | 1490 | strong social studies (as measured by the percentage of students majoring in social studies  > 30 percent) | 
| Cost-conscious Casey | 1600 | wants a public school in CA or a school where students from homes in the bottom 20th percentile of incomes pay less than 10000 per year|

Here are the rules. They want to go to school where their test scores are within 100 points of the school average SAT score. To match their preferences, use the most recent data. You will need a few tools.

1. First, in order to understand what a column contains, you can use the `distinct()` function^[from `dplyr`. The codebook is also useful.]. For example, say you are trying to figure out how to identify "ivy plus" schools (or what that specifically means). Notice that there is a columns called  `tier_name`, then run the code (don't worry if you are not familiar with `%>%` at this moment. We will cover this in Lecture 7):

    ```{r}
    mrc_data %>% distinct(tier_name)    
    ```
    We see there are 12 tiers and one is "Ivy Plus"! Note the capitalization.
    
1. Second, we're going to have to find schools that match ranges of SAT scores by using `filter()` function, which filters out all observations that meet the given condition. We can use the `between()` function from `dplyr`. Are the following two approaches generating the same output? If not, what is the difference?

    ```{r, eval = FALSE}
    mrc_data %>% filter(1330 < sat_avg_2013, sat_avg_2013 < 1530)
    mrc_data %>% filter(between(sat_avg_2013, 1330, 1530))
    ```

*Solution: They are different. between() includes the end points (i.e. it uses <=)*

1. The final thing is a concept. You're probably about to write code that looks like the following pseudo code.^[pseudo code is a term for fake code that captures the logic of some coding idea without being actual code.] 

1. Now you're ready to find the college for the five friends.

    ```{r, eval = FALSE, echo = FALSE}
    # fill in the  ... with appropriate code
    
    # We'll give this a name so we can use it later.
    bff_super_awesome_college_list <- 
    mrc_data %>%
      mutate(abdul_choices = ifelse(between(sat_avg_2013, 1330, 1530) & 
                                    (tier_name == "Ivy Plus" | ... ), TRUE, FALSE),
             sam_choices = ifelse(..., ..., ...), 
             nancy_choices = ifelse(..., ..., ...),
             rei_choices = ifelse(..., ..., ...),
             casey_choices = ifelse(..., ..., ...)
             ) 
      
    
    bff_super_awesome_college_list %>%
      filter(abdul_choices, sam_choices, nancy_choices, rei_choices, cary_choices)
    ```

  a. What school(s) are acceptable to all five?
  a. How many school(s) are available to any of the five? Adjust `filter` statement slightly.^[Hint: Think about the warm-up you did for this lab]
  <!-- Note to Reviewer ... any suggestion to make this clearer would help -->

```{r, eval = FALSE, echo = TRUE}
# SOLUTION
 bff_super_awesome_college_list <-
    mrc_data %>%
      mutate(abdul_choices = ifelse(between(sat_avg_2013, 1330, 1530) & 
                                  (tier_name == "Ivy Plus" | flagship == 1 ), 
                                  TRUE, FALSE),
            casey_choices = ifelse(between(sat_avg_2013, 1500, 1600) &
                                  (public == 1 & state == "CA" | 
                                  scorecard_netprice_2013 < 10000), 
                                  TRUE, FALSE),
            rei_choices = ifelse(between(sat_avg_2013, 1390, 1590) &
                                (pct_socialscience_2000 > 30), 
                                 TRUE, FALSE),
            nancy_choices = ifelse(between(sat_avg_2013, 1490, 1600) &
                                  (region == 2), 
                                  TRUE, FALSE),
            sam_choices = ifelse(between(sat_avg_2013, 1350, 1550) &
                                 (public == 0), 
                                 TRUE, FALSE),
             ) 
      
bff_super_awesome_college_list %>%
  filter(abdul_choices, casey_choices, rei_choices, nancy_choices, sam_choices)

```

1. The five friends have `NA` in their choice sets. Do the the school list change if we replace all the `NA`s with `TRUE`? Without coding, argue why the list will not change if we replace the `NA`s with `FALSE`.

*SOLUTION:* When we make force missing values to be `TRUE`, they still only agree on UChicago. If we replace the `NA` with `FALSE` the results will not change following the logic from the warm-up. In filter, we're asking a bunch of logical AND (`&`) so if we replace an `NA` with `FALSE` we'll continue to filter away that row!


```{r, eval = FALSE}
# The straight forward way to solve is to fill all the NAs with TRUE using a 
# new call to ifelse() for each person
# ex: 
bff_super_awesome_college_list %>%
  mutate(abdul_choices = ifelse(is.na(abdul_choices), TRUE, abdul_choices),
         casey_choices = ifelse(is.na(casey_choices), TRUE, casey_choices),
         rei_choices = ifelse(is.na(rei_choices), TRUE, rei_choices),
         nancy_choices = ifelse(is.na(nancy_choices), TRUE, nancy_choices),
         sam_choices = ifelse(is.na(sam_choices), TRUE, sam_choices)) %>%
filter(abdul_choices, casey_choices, rei_choices, nancy_choices, sam_choices)


# you might Google around and figure out something like this; though tbh it
# took a while for me to find it.
bff_super_awesome_college_list %>%
  replace_na(list(abdul_choices = TRUE, 
                  casey_choices = TRUE, 
                  rei_choices = TRUE, 
                  nancy_choices = TRUE, 
                  sam_choices = TRUE)) %>%
  filter(abdul_choices, casey_choices, rei_choices, nancy_choices, sam_choices)

  

# Another approach is to use if_else() which is a dplyr ifelse() that has 
# accepts a missing argument.

    mrc_data %>%
      mutate(abdul_choices = if_else(between(sat_avg_2013, 1330, 1530) & 
                                    (tier_name == "Ivy Plus" | flagship == 1 ), 
                                    TRUE, FALSE, missing = TRUE),
            casey_choices = if_else(between(sat_avg_2013, 1500, 1600) &
                                    (public == 1 & state == "CA" | 
                                       scorecard_netprice_2013 < 10000), 
                                    TRUE, FALSE, missing = TRUE),
            rei_choices = if_else(between(sat_avg_2013, 1390, 1590) &
                                   (pct_socialscience_2000 > 30),
                                    TRUE, FALSE, missing = TRUE),
            nancy_choices = if_else(between(sat_avg_2013, 1490, 1600) &
                                     (region == 2), 
                                    TRUE, FALSE, missing = TRUE),
            sam_choices = if_else(between(sat_avg_2013, 1350, 1550) &
                                          (public == 0), 
                                  TRUE, FALSE, missing = TRUE),
             ) %>%
  filter(abdul_choices, casey_choices, rei_choices, nancy_choices, sam_choices)


```



1. **Challenge (optional)** Create a "Five friends college ranking". A college is ranked 1 if it is acceptable to all 5 friends. 2 if it is acceptable to any 4 friends and so on.^[3 if it is acceptable to 3 friends. 4 if acceptable to 2 friends and 5 if acceptable to 1 friend] Colleges that are not acceptable to any friend should be marked "Unranked".
