---
title: "Lab 1 Solved: Introduction to R"
author: "Harris Coding Camp"
date: "Summer 2023"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
```


Open an R script so you can save your code for later! 


1. If you have not yet, run the following code to install the `tidyverse`.

    ```{r, eval = FALSE}
    # download from the internet -- you only need to do this once.
    install.packages("tidyverse")
    ```

2. This code returns an error. Why? Fix it and install the packages. 


    *Answer: need to put package names in quotes here. `install.packages("haven")`*

    ```{r, eval = FALSE}
    
    install.packages(haven)
    install.packages(readxl)
    ```

3. In order to have access to `tidyverse` functions, you need to load the library. Run the following code.

    ```{r, eval = FALSE}
    library(tidyverse)
    ```

4. Run the following code to make sure you have installed successfully. Then, use the function `View()` to see the full dataset. 

    ```{r, eval = FALSE}
    # storms is a dataset that comes with the tidyverse
    # use View(storms) to see the full dataset
    big_storms <- storms %>%
      group_by(name, year) %>%
      filter(max(category, na.rm = TRUE) == 5)
    ```

5. Assign the code above to the name `big_storms`. What is the difference between `storms` and `big_storms`?

    *big_storms has fewer storms, because we remove storms that never made it category 5*

6. This code makes the map seen below. If you run it you will get a message that says your code is missing a required package, R 4.1 will ask you to install the package. Choose yes. [This package is not essential -- if the package fails to install, make a note of what error messages you get and then move on. You may consult with TAs after finishing the tasks below.]

```{r, eval= FALSE}
  # Heads up: copy and paste might mess up the quotes or other formatting
  # look carefully if the code doesn't work.
  ggplot(aes(x = long, y = lat, color = name), data = big_storms) +
        geom_path() + 
        borders("world") +  
        coord_quickmap(xlim = c(-130, -60), ylim = c(20, 50))
```

```{r, echo = FALSE}
    storms %>%
      group_by(name, year) %>%
      filter(max(category) == 5) %>%
  ggplot(aes(x = long, y = lat, color = name)) +
        geom_path() + 
        borders("world") + 
        coord_quickmap(xlim = c(-130, -60), ylim = c(20, 50))
    
```

\newpage
 
## Warm-up

1. Which of these allow you to pull up the documentation for a command in R?

  a. `*`
  *b. `?`*
  c. `help()`
  d. `documentation()`

2. In the code block below, run code that will pull up documentation for the function `paste0()`.
```{r, eval= FALSE}
?paste0()
```

What does this function do?

*paste0() concatenates vectors and makes them characters*

3. The second example in ?paste0 is

```{r}
## If you pass several vectors to paste0, they are concatenated in a
## vectorized way.
nth <- paste0(1:12, c("st", "nd", "rd", rep("th", 9)))
```

This example uses a bunch of code/concepts we haven't covered yet! 

a. Try to make sense of how the code works by running each input seperately. i.e. what does the code `1:12`, `c("st", "nd", "rd", rep("th", 9))` and `rep("th", 9)` do? And what does `paste0` do with it's inputs?
b. The function `paste0` is given two inputs here. What are they?
    *1:12 is a vector of numbers and c("st", "nd", "rd", rep("th", 9)) is a vector of charatcters*
c. What does `paste0` do with two vector inputs?
    *concatenates them in a vectorized way*

4. *Assigning variables.*  `score` is tracking your score in a game. The first round you got 3 points, the next round you got 2 points.


Your partner is keeping track of your score and wrote the following code.

```{r, eval = FALSE}
# Solution
score <- 0
# round 1
score <- score + 3
# round 2
score <- score + 2
```

a. What is `score` now? 
    
    *score was still 0*
  
b. Fix the code so that `score` is accurately tracking your score.

Comment:  A common error for new programmers is to not assign output to a name!

## The rest of the problems are answered in the lab.