---
title: 'Accelerated Lecture 4: If statements and conditionals'
author: "Harris Coding Camp -- Standard Track"
date: "Summer 2022"
output:
  beamer_presentation:
    fig_width: 6
    fig_height: 4
    includes:
      in_header: ../resources/header_page_n.tex
---
```{r, setup=TRUE, include=FALSE}
library(tidyverse)
library(readr)
options(dplyr.summarise.inform = FALSE) # Suppress summarize info in dplyr >= 1.0

south_africa_data <- readxl::read_xlsx("../../labs/Data/world_wealth_inequality.xlsx",
                         col_names = c("country", "indicator", "percentile", "year", "value")) |>
  separate(indicator, sep = "[\\r]?\\n",
          into = c("row_tag", "type", "notes", "empty")) |> 
  filter(country == "South Africa", !is.na(value)) |>
  select(country, percentile, year, value)
```

## Review: Subsetting data

```{r, eval = FALSE}
# tidyverse
data |>
  filter(row_condition) |>
  select(columns, we, want)

# base R
data[row_condition, c("columns", "we", "want"]
```


## Review: Do these `filter()` calls give the same result?

```{r, eval = FALSE}
south_africa_data |>
  filter(percentile == "p0p50",
         value == max(value, na.rm = TRUE))

south_africa_data |>
  filter(percentile == "p0p50") |>
  filter(value == max(value, na.rm = TRUE)) 
```

## Review: Why not?

```{r, eval = TRUE}
identical(
south_africa_data |>
  filter(percentile == "p0p50",
         value == max(value, na.rm = TRUE)),
south_africa_data |>
  filter(percentile == "p0p50") |>
  filter(value == max(value, na.rm = TRUE)) 
)
```


## Review: Subsetting data


```{r, eval = FALSE}
south_africa_data |>
  filter(percentile == "p0p50",
         value == max(value, na.rm = TRUE))

# this is the max for ALL income groups.
max(south_africa_data$value, na.rm = TRUE) 


```

## Review: Subsetting data
\footnotesize
Here, we use the `max()` on the bottom half only.

```{r, eval = FALSE}
south_africa_data |>
  filter(percentile == "p0p50") |>
  filter(value == max(value, na.rm = TRUE)) 
```

We could rewrite the code like so.

```{r}
bottom_half <- south_africa_data |>
  filter(percentile == "p0p50") 

# this is the max for the bottom half income groups.
bottom_half |> 
  filter(value == max(value, na.rm = TRUE)) 
```


## Review: Sorting data

```{r, eval = FALSE}
# tidyverse
arrange(data, col, desc(col2))

# base R
data[order(data$col, -data$col2),]
```


## Review: Summarizing data

```{r, eval = FALSE}

# tidyverse 
# results in a tibble with 1 row
summarize(data, mean = mean(col))

# base R
# results in a vector of length 1
mean(data$col)
```

\footnotesize
- Let `mean` stand in for any function that *reduces* a vector to a **single value**

## Review: Creating new data


```{r, eval = FALSE}
# tidyverse
data <- data |> mutate(new_column = something) 

# base R
data$new_column <- something 
```


Same functionality to change old data:

```{r, eval = FALSE}
# tidyverse
data <- data |> mutate(old_column = something) 

# base R
data$old_column <- something 
```

\footnotesize
- `something` is a vector length nrow(data) or 1

# Quiz time -- no stakes!!

##  Do your best! Do it by yourself!

- Only resource is R and RStudio


Go to Canvas

- Click on `Gradescope`
- Click on `Accelerated Quiz`
- 10 minute timer starts when you open it




## How would we make a column dependent on other data?
\footnotesize
```{r, echo = FALSE}
set.seed(12)
d <- tibble(x = 1:4,
       y = round(rnorm(4), 1))

d
```

Add column dependent on `y`
```{r, echo = FALSE}
d |> mutate(set_neg_y_to_0 = ifelse(y < 0, 0, y))
```


# Call in `if` and `ifelse`


## When we want code to do something depending on the context


```{r eval=FALSE}

ifelse(test, yes, no)


if (test is TRUE) {
  do this
} else {  # test is FALSE                    
  do this other thing
}
```

We will cover:

 - introduce vectorized `ifelse` and `case_when()` statements 
 - introduce `if` and `else` statements



##  Syntax: `ifelse(test, yes, no)` 

- **if**`test` is TRUE return  `yes`
- **else** `test` is FALSE return `no`

```{r}
x <- c(5, 50)

ifelse(x > 10, "x is big", "x is small")
```


##  `ifelse(test, yes, no)` is vectorized

```{r}
x <- c(5, 50, log(1e11), -1)
# 5 > 10 ?          ... no
# 50 > 10 ?         ... yes
# log(1e11) > 10 ?  ... yes
# -1 > 10 ?         ... no
ifelse(x > 10, "x is big", "x is small")
```

## What will the following statements return? 

```{r eval=FALSE}

ifelse(is.na(NA), 1, 2)


ifelse(is.na("a"), 1, 2)
```

## `test` typically evaluates to a boolean vector 

Think of:

- conditional operators
- is.<type>() tests

```{r eval=TRUE}
# ifelse(is.na(NA), 1, 2)
ifelse(TRUE, 1, 2)

# ifelse(is.na("a"), 1, 2)
ifelse(FALSE, 1, 2)
```


## `TRUE` gives option 1, `FALSE` gives option 2

```{r eval=TRUE}
# ifelse(TRUE, 1, 2)
ifelse(is.na(NA), 1, 2)


# ifelse(FALSE, 1, 2)
ifelse(is.na("a"), 1, 2)
```


## What will the following statements return? 

```{r eval=FALSE}
ifelse(c(TRUE, FALSE, FALSE, TRUE), "a", "b")

ifelse(1:4 > 3, "a", "b")

```

## What will the following statements return? 

```{r eval=TRUE}
ifelse(c(TRUE, FALSE, FALSE, TRUE), "a", "b")

ifelse(1:4 > 3, "a", "b")
```

## Another example

```{r}
trial_1 <- c(98, 20, 100, 18, 40)
trial_2 <- c(30, 41, 64, 8, 70)

ifelse(trial_1 > trial_2, trial_1, trial_2)
```


## What should the following code returns?

```{r, eval = FALSE}
states <- c("IL", NA, "IA", "NM")

ifelse(states == "IL", "home", "elsewhere")
```

## `NA` still contagious

```{r}
states <- c("IL", NA, "IA", "NM")

ifelse(states == "IL", "home", "elsewhere")
```



## Using `ifelse` with data
\footnotesize
Add a column called vowel which is 1 for "a", "e", "i", "o" and "u" and 0 otherwise.^[Sorry "y"!]

```{r}
alphabet <- tibble(letters = letters)
head(alphabet)
```

## Call `ifelse()` inside `mutate()`!


\footnotesize
```{r}
alphabet |>
  mutate(vowel = 
          ifelse(letters %in% c("a", "e", "i", "o","u"), 
                 1, 0)) |>
  head()
```

Code that works on vectors, will work on columns. After all, they're vectors!

## Of course, you'll see baseR do this too

\footnotesize

```{r, eval = FALSE}
alphabet$vowel <- 
  ifelse(alphabet$letters %in% c("a", "e", "i", "o","u"), 
         1, 0)
```

# Are we stuck with two outcomes?



## `ifelse` statements with multiple categories

Let's make the `vowel` column 

- "yes" for "aeiou"
- "sometimes" for "y" 
- "no" for everything else

```{r}
tail(alphabet)
```

## Option 1: call `ifelse` multiple times

```{r}
alphabet |>
  mutate(
    vowel = "no",
    vowel = ifelse(letters == "y", "sometimes", vowel),
    vowel = ifelse(letters %in%  c("a", "e", "i", "o", "u"), 
                   "yes", vowel)
  ) |>
  tail()
```
## Option 1: call `ifelse` multiple times

```{r}
alphabet |>
  mutate(
    vowel = "no"
  ) |>
  tail()
```


## Option 1: call `ifelse` multiple times

```{r}
alphabet |>
  mutate(
    vowel = "no",
    vowel = ifelse(letters == "y", "sometimes", vowel)
  ) |>
  tail()
```



## Option 1: call `ifelse` multiple times

```{r}
alphabet |>
  mutate(
    vowel = "no", 
    vowel = ifelse(letters == "y", "sometimes", vowel),
    vowel = ifelse(letters %in%  c("a", "e", "i", "o", "u"), 
                   "yes", vowel)
  ) |>
  tail()
```


## Option 2: nest the `ifelse`

```{r}
alphabet |>
  mutate(
    vowel = ifelse(letters == "y", 
                   "sometimes", 
            ifelse(letters %in%  c("a", "e", "i", "o", "u"),
                   "yes", "no"))
  ) |>
  tail()
```

## option 3: `case_when()`

```{r}
alphabet |>
  mutate(
    vowel = 
      case_when(
        letters == "y" ~ "sometimes", 
        letters %in%  c("a", "e", "i", "o", "u") ~ "yes",
        TRUE ~  "no")
  ) |>
  tail()
```


## Another nested `ifelse` example
\footnotesize
```{r}
txhousing |> 
  select(city, year, month, median) |>
  mutate(housing_market = 
      ifelse(median < 100000, "first quartile",
      ifelse(median < 123800, "second quartile",
      ifelse(median < 150000, "third quartile",
      ifelse(median < 350000, "fourth quartile",
             NA))))
           ) |> 
  head(3)
```


## `case_when` again
\footnotesize

```{r eval=FALSE}
# add a column called `housing_market` to the `txhousing` 
txhousing |> 
  select(city, year, month, median) |>
  mutate(housing_market = 
         case_when(
           median < 100000 ~ "first quartile",
           median < 123800 ~ "second quartile",
           median < 150000 ~ "third quartile",
           median < 350000 ~ "fourth quartile"
         ))  |>
  head(3)
```


## `case_when` statements are a bit "surly"
\footnotesize
`case_when` will not do type coercion. 

```{r, eval = FALSE}
txhousing |> 
  mutate(housing_market = 
        case_when(
           median < 100000 ~ 1,
           median < 123800 ~ "second quartile",
           median < 150000 ~ "third quartile",
           median < 350000 ~ "fourth quartile"
         )) |> 
  select(city, median, housing_market)

Error: must be a double vector, not a character vector
Run `rlang::last_error()` to see where the error occurred.
```

Here we try to include *both* doubles and characters in the `housing_market` column, but atomic vectors can only have one type!

  - Rather than coerce and provide a warning, the developers decided to make this an error
  - If using `NA` as an output, you have to specify `NA` types e.g. `NA_integer_`, `NA_character_`
  
  
  
## `case_when` "else"
\footnotesize
You might wonder how to approximate `else`. 

 - Use `TRUE` as a catch all. 

```{r, eval = FALSE}
example <- tibble(a = 1:12)

example |>
  mutate(category = case_when(a %in% c(2, 3, 5, 7, 11) ~ "prime",
                              sqrt(a) == round(sqrt(a))  ~ "square",
                              TRUE ~ "other"))
```


## Try it yourself
\footnotesize
We will use `midwest` here, which is a dataset built into `tidyverse`.

1. Create a new variable, `poverty_designation`, that is "High Poverty" if `percbelowpoverty` is above 10 and is "Low Poverty" otherwise.

```{r, eval=FALSE, echo = FALSE}
# answer
midwest$poverty_designation <- ifelse(midwest$percbelowpoverty > 10, "High Poverty", "Low Poverty") 
```

2. Create a new variable called `ohio` that is "Ohio Counties" for observations from Ohio and "Other Midwestern Counties" for the rest of the observations.

```{r, eval=FALSE, echo = FALSE}
# answer
midwest$ohio <- ifelse(midwest$state == "OH", "Ohio Counties", "Other Midwestern Counties") 
```

3. Create a new variable called `populous_counties` that is `TRUE` for the observations from the counties listed in `big_counties` and `FALSE` otherwise. \footnotesize
```{r, eval = FALSE}
big_counties <- c("COOK", "WAYNE", "CUYAHOGA", "OAKLAND", "FRANKLIN")
```

```{r, eval=FALSE, echo = FALSE}
# answer
big_counties <- c("COOK", "WAYNE", "CUYAHOGA", "OAKLAND", "FRANKLIN")
midwest$populous_counties <- midwest$county %in% big_counties
```

4. Create a new variable called `pop_index` that is "High" for the observations with `poptotal` greater than 100000, is "Medium" for the observations with `poptotal` between 30000 and 100000, and "Low" otherwise.

```{r, eval=FALSE, echo = FALSE}
# answer
midwest$pop_index <- 
  ifelse(midwest$poptotal > 100000, "High", 
    ifelse(midwest$poptotal >= 30000 & midwest$poptotal <= 100000, "Medium", 
      "Low")
```

# `if` statements

## `if` statements
 

```{r eval=FALSE}
if (condition is TRUE) {
  do this
  ...
  ...
  ...
}
```



## `if` statements

For example:
```{r}
x <- 100

if (x > 0) {
  print("x is positive")
}

```

## `if/else` statements



```{r eval=FALSE}
if (condition is TRUE) {
  do this
} else {                      
  do this other thing
}
```

## `if/else` statements, example

```{r}
x <- -5
if (x > 0) {
  print("Non-negative number")
} else {
  print("Negative number")
}
```


## `if` and `else` versus `ifelse`


`ifelse` 

  - often used in a data setting
  - handy for quick `yes`, `no` type alternatives
  - vectorized and accepts `NA`
  
`if` and `else` 
  
  - often used in a "programming" setting
  - handle complicated chunks of code and more complex alternatives
  - `if()` only accepts TRUE or FALSE (not vectorized, no `NA`)


## `if`, `else if` and `else` statements

If we have more than 2 conditions, use `if`, `else if` and `else`:

```{r eval=FALSE}
if (condition is TRUE) {
  do this
} else if (second condition is TRUE) {                      
  do this other thing
} else if (third condition is TRUE) {
  do this third thing
} else {
  do a default behavior
}
```

Note: a default behavior with `else` is not necessary.
 


## `if`, `else if` and `else` statements, example

\footnotesize
```{r}
x <- sample(1:100, 1)
x
y <- sample(1:100, 1)
y
```

\footnotesize
```{r}
if (x > y) {
  print("x is greater")
} else if (x < y) {
  print("y is greater")
} else {                
  print("x and y are equal")
}
```

## `if`, `else if` and `else` can take a compound condition

```{r}
x <- sample(1:100, 1)
x
y <- sample(1:100, 1)
y
z <- sample(1:100, 1)
z
```

## `if`, `else if` and `else` can take a compound condition

```{r}
if (x >= y & x >= z) {
  print("x is the greatest")
} else if (y >= z) {
  print("y is the greatest")
} else {                
  print("z is the greatest")
}
```

## Try it yourself
\footnotesize
Let's develop a small dice game. 

1. Fill in the ... so the code says "You win" if the dice add up to 7 and "You lose" otherwise.

```{r, eval = FALSE}
dice <- sample(c(1:6), 2)
    
if (...) {
  print("You win")
} else {
  print("You lose")
}
```

2. Add an `else if()` block to the code above that says "try again" if the dice add up to 6 or 8.


## Try it yourself
\footnotesize
2. Add an `else if()` block to the code above that says "Try again" if the dice add up to 6 or 8.

```{r, eval = FALSE}
dice <- sample(c(1:6), 2)
    
if (...) {
  print("You win")
} else if (...) {
  print("Try again")
} else {
  print("You lose")
}

```

```{r, eval = FALSE, echo=FALSE}
# answer
dice <- sample(c(1:6), 2)

if (sum(dice) == 7) {
  print("You win")
} else if(sum(dice) %in% c(6,8)) {
  print ("Try again")
} else {
  print("You lose")
}

```


## Some common uses of `if`

Sharing code among various people. 

- `Sys.getenv("USER")` returns the name of the USER fr

```{r, eval = FALSE}

if (Sys.getenv("USER") == "arianisfeld") {
  setwd("~/repo/dir")
} else if (Sys.getenv("USER") == "yunjoo") {
  setwd("C://repo/dir")
} else {
  print(paste0("WARNING: Unknown user. 
               Working directory is ", getwd()))
}

```

## `if()` the condition must return TRUE or FALSE

`if()` is not vectorized

```{r, eval = FALSE}
x <- c(1, -4)

if (x > 0) {
   x
} else {
  -x
}
```

`Error in if (x > 0) { : the condition has length > 1`


## `if()` the condition must return TRUE or FALSE


`if()` does not handle `NA`s

```{r, eval = FALSE}
x <- NA

if (x > 0) {
  x
} else {
  -x
}
```

`Error in if (x > 0) { : missing value where TRUE/FALSE needed`


## If you can't afford errors ...

write code to handle edge cases

```{r, eval = TRUE}
x <- NA
if (length(x) == 1 & all(!is.na(x) & x > 0)) {
  x
} else if (length(x) == 1) {
  -x
}


x <- c(pi, 2)
out <- if(length(x) == 1 & all(!is.na(x) & x > 0)) {
  x
} else if(length(x) == 1) {
  -x
}
out
```

## Detour: Why `NULL`? Why not `NA`?
\footnotesize
`NULL` stands in for an *object* that is undefined.

```{r}
length(NULL)
NULL > 1
```  

~~~

`NA` stands in for a *value* that is undefined.

```{r}
length(NA)
NA > 1
```  



## `if()` the condition must return TRUE or FALSE
\footnotesize
Good idea to make sure it still works for valid input!

```{r}
x <- exp(1)
if (length(x) == 1 & all(!is.na(x) & x > 0)) {
  x
} else if (length(x) == 1) {
  -x
} 


x <- -1000
if (length(x) == 1 & all(!is.na(x) & x > 0)) {
  x
} else if (length(x) == 1) {
  -x
} 

```




## Recap

Today we learned how to:

- use control flow with `if` and `ifelse` statements
- use `ifelse()` and `case_when()` statements in conjunction with `mutate()` or `$<-` to create columns based on conditional statements

## Next up

Lab:

- Today: Practice with `ifelse`

**I can use `ifelse` to create columns conditional on data**


Lecture:

- Making data visualizations

