---
title: "Flash Cards"
author: "Harris Coding Camp"
date: "Summer 2023"
output:
  beamer_presentation:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)

```

# Name three functions for loading data into R.

## Name three functions for loading data into R.

```{r, eval= FALSE}
readr::read_csv()  # preferred over base::read.csv()
haven::read_dta()
readxl::read_xlsx()
```

You'll learn about more depending on your needs.


# How do you test if a value is *not* missing?

##  How do you test if a value is *not* missing?

```{r}
!is.na(c(1, NA, 100))
```



# Take a vector `c('anisfeld', 'mkr')`. What function let's you add `'@uchicago.edu'` to each item?

## What function do you use to add `'@uchicago.edu'` to each item?

`paste0()` or `str_c()` 

```{r}
cnet <- c('anisfeld', 'mkr')

paste0(cnet, "@uchicago.edu")
# tidyverse function in package stringr
str_c(cnet, "@uchicago.edu")
```

# What does this code return? `max(c(100, -10, NA))` 

## What does this code return? `max(c(100, -10, NA))` 

```{r}
max(c(100, -10, NA))

# if it's safe to ignore NA
max(c(100, -10, NA), na.rm=FALSE)
```



# Gotchas



## Your friend gets the following error. What's the culprit? 


```{r, eval = FALSE}
read_dta("my_file.dta")
```
Error in read_dta("my_file.dta") : could not find function "read_dta"

## What's the culprit? 


They need to load the `haven` package

```{r, eval = FALSE}
library(haven)
read_dta("my_file.dta")

# or use double colons
haven::read_dta("my_file.dta")
```



## Your friend gets the following error. What's the culprit? 


```{r, eval = FALSE}
read_xlsx("data/my_file.xlsx")
```
Error: `path` does not exist: ‘data/my_file.dta’


## What's the culprit? 


1. There's a typo in the file name.
2. The file or folder is not in the working directory.


## Your friend gets the following error. What's the culprit? 

```{r, eval = FALSE}
install.packages(bench)
```
Error in install.packages : object 'bench' not found


##  What's the culprit? 

The packages need to be a *character* vector of the names of packages (i.e. in quotes)

```{r, eval = FALSE}
install.packages("bench")
```


When they use `bench` without quotes, R looks for it in the global environment! 


