---
title: "Lab 4 Solved: base R with vectors and data frames"
author: "Harris Coding Camp"
date: "Summer 2023"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
midwest_data <- midwest
```

# Using `[` and other base R tools for data analysis.

## Warm-up

We’ll use midwestern demographic data which is at this [link](https://github.com/harris-coding-lab/harris-coding-lab.github.io/raw/master/data/midwest.dta). The dataset includes county level data for a single year. We call data this type of data “cross-sectional” since it gives a point-in-time cross-section of the counties of the midwest.

```{r}
# SOLUTION
midwest_data <- haven::read_dta("../Data/midwest.dta")
nrow(midwest_data)
ncol(midwest_data)
names(midwest_data)
```


1. What format^[i.e. is it a csv, dta, xlsx?] is the midwest data in? What function do you need to load it?
2. Load the package so you can read in the data and assign it to the name `midwest_data`. If you don't remember what package contains that function use `??`  (as in `??read_xxx`) 

    `??` is a way to search through help for functions that are not currently loaded in R or if you forgot the exact name of a function.
    
3. How many rows and columns does the data have? 
4. Notice that row represents a county which is uniquely identified by a `PID` or using `county` + `state`.
5. Use `names()` to see the names of all the columns. 


# Using `[` and `$` with vectors.

Recall that columns are vectors and you can extract the vector using `$`.

1. Extract the `inmetro` and calculate the mean. 
    
    ```{r, include = TRUE}
    # SOLUTION
    mean(midwest_data$inmetro)
    ```
    
    We might interpret the result as the proportion of counties that are urban ... but the data did not come with a codebook, so we are not sure! Let's explore.

1. Run the following code and explain in words what the two results are. Assign the two resulting vectors to names that reflect what they are.
    
    ```{r, eval = FALSE}
    midwest_data$poptotal[midwest_data$inmetro == 1] 
    midwest_data$poptotal[midwest_data$inmetro == 0]
    ```


1. What is the average (mean) population for urban midwest counties? How about non-urban counties? Do these numbers make sense?
1. What are the `max()` and `min()` for these vectors? Do these numbers make sense?
1. How many "urban counties" have fewer than 50000 residents. What proportion of the counties is this?


```{r, include = TRUE}
# SOLUTIONS
urban_pop <- midwest_data$poptotal[midwest_data$inmetro == 1] 
rural_pop <- midwest_data$poptotal[midwest_data$inmetro == 0]

mean(urban_pop)
mean(rural_pop)

min(urban_pop)
max(urban_pop)

min(rural_pop)
max(rural_pop)


length(urban_pop[urban_pop < 50000])
length(urban_pop[urban_pop < 50000])/length(urban_pop)

# this is a bit more sophisticated
sum(urban_pop < 50000)
mean(urban_pop < 5000)

```

1. You can use `sort()` to see the numbers in order.
    
1. What is the population of the 10th smallest urban county in the midwest? How about the 10th largest? (Hint: Use `?sort` to learn how to change the order of your sort.)

    ```{r, echo = TRUE}
    # SOLUTION
    # 10th smallest
    sort(midwest_data$poptotal[midwest_data$inmetro == 1] )[10]
    # 10th largest
    sort(midwest_data$poptotal[midwest_data$inmetro == 1], decreasing = TRUE )[10]
    ```

1. What are the name of the 4 urban counties with population under 20000? (You can use `&` to combine conditional expressions.)
1. What are the PID of the 4 urban counties with population under 20000?

```{r, include=TRUE}
# SOLUTION
midwest_data$county[midwest_data$poptotal < 20000 & midwest_data$inmetro == 1]
midwest_data$PID[midwest_data$poptotal < 20000 & midwest_data$inmetro == 1]
# we could also use double indexing as we discuss below.
midwest_data[midwest_data$poptotal < 20000 & midwest_data$inmetro == 1, "PID"]
midwest_data[midwest_data$poptotal < 20000 & midwest_data$inmetro == 1, "state"]
```
1. What states are those counties in?

*See above*


1. *Common error*: Run the code below and get an error that says "Column `County` doesn't exist". Fix the error.

```{r, eval=FALSE}
midwest_data[midwest_data$poptotal < 20000 & midwest_data$inmetro == 1, 
             c("County", "state", "poptotal", "popdensity", "inmetro")]
```

*This is a typo, the column name is `country`*

1. Adjust the code above so we get the same rows, but see all the columns.

```{r, include = TRUE}
midwest_data[midwest_data$poptotal < 20000 & midwest_data$inmetro == 1, ]
```

1. These are the low population counties where `inmetro` is 1.  Are the population densities low? Compare them to the population densities of similar population counties where `inmetro` is `0`. If you have time, you can look at the locations on a map and get a better understanding of what `inmetro` captures.

```{r, include = TRUE}
# the popdensity are similar for these regions, though the inmetro == 1 counties 
# are slightly bigger
midwest_data[midwest_data$inmetro == 0 & midwest_data$poptotal < 20000, ]

midwest_data[midwest_data$inmetro == 0,  "popdensity"]
```

**Rapid fire:** 

Using `[` and `$` complete the following challenges:

1. What states have an Adams County?
1. How many counties are in Indiana?
1. What county has the highest percent Asian in this data? 
1. Make a data frame that includes the 10 largest counties (by total population) and shows the county, state, total population, and percent with college degree. Assign the output to the name `top_ten`.


```{r, include = FALSE}
midwest_data[midwest_data$county == "ADAMS", "state"]
nrow(midwest_data[midwest_data$state == "IN", ])
midwest_data[midwest_data$percasian == max(midwest_data$percasian), ]

tenth_largest <- sort(midwest_data$poptotal, decreasing = TRUE)[10]
top_ten <- midwest_data[midwest_data$poptotal >= tenth_largest, c("county", "state", "poptotal", "percollege") ]

```


1. A student tried to answer "How many counties are in Indiana?" with the following code:

```{r}
length(midwest_data[midwest_data$state == "IN", ])
```

What does `length()` count for `tibble`s and `data.frame`s?^[Use `nrow()` instead!]

*Length provides the number of columns! This makes sense if we keep in mind that the tibble is a list and the relevant length is the number of "pointers" need to store the list.*

## Ordering columns

It would be nice to sort the table with the 10 largest counties by population. How do we do that?

Let's start with a simpler example -- this is a good way to get intuition for what the code does! 


1. Create the following test data set. Your random numbers will be different!

    ```{r}
    test_data <- tibble(
      id = c(1, 4, 2, 3, 5),
      gpa = 4 * runif(5)
    )
    
    test_data
    ```

1. Explain what the following code does.

    ```{r}
    test_data[c(1, 2),]
    test_data[c(2, 1),]
    ```
    Did you notice that the order of the rows changed! This suggests that we could re-order or sort the data frame, if we knew the correct **order** of the rows.
    
1. Pull out the rows that correspond to ids 1, 2, 3.  (e.g. the **3**rd row corresponds to `id` number 2). To make a data frame like this:

    ```{r, echo = TRUE}
    # Solution
    test_data[c(1,3,4), ] 
    ```

1. base R has a function called `order()` which tells you the order of the rows (in increasing order). Use `order()` on the test_data ids. Plug this into your `[` to sort the data.

    ```{r, echo = TRUE}
    # Solution
    test_data[order(test_data$id), ] 
    ```

1. Now sort `top_ten` in **decreasing** order. The expected output is:

    ```{r, echo = TRUE}
    # Solution
    top_ten[order(top_ten$poptotal, decreasing = TRUE), ] 
    ```

## II. Investigate the `diamonds` dataset

Throughout this exercise, we will be working with the `diamonds` dataset (comes with `tidyverse`), which contains the prices and other attributes of almost 54,000 diamonds. (use `?diamonds` to see the codebook.)

1. Run the following command to familiarize ourselves with this dataset. How many observations and variables are included in `diamonds`? 

    **SOLUTION: 53,940 observations (rows) and 10 variables (columns)**
    
```{r, eval = FALSE}
# tidyverse
glimpse(diamonds)
# base R (utils)
str(diamonds)
```

2. Try to describe the shape and center of the `price` distribution by observing the summary statistics like the mean, median and quartiles.

  **SOLUTION: We see that the mean is much larger than the median, which suggests there are some very expensive diamonds in the right tail of the distribution.**

```{r, eval = FALSE}
summary(diamonds$price)
```

3. How many diamonds cost less than \$500? less than \$250? How many diamonds cost \$15000 or more?

    **Solution: from the summary, we see there are no diamonds less than \$250. It's also clear that <\$500 and >\$15000 are a small portion of the data as both values are close to the extremes and outside of the middle 50 percent of the data (aka the interquartile range). Specifcally:**
    
```{r}
nrow(diamonds[diamonds$price < 500, ])    
nrow(diamonds[diamonds$price > 15000, ])    

```
    
4. Which cut has the highest priced diamond? What is the price?

```{r}
# bracket and dollar sign solution 
diamonds[diamonds$price == max(diamonds$price), c("cut", "price")]
```


5. Redo part 4 with the lowest priced diamond.

```{r}
diamonds[diamonds$price == min(diamonds$price), c("cut", "price")]
```

6. Is there any relationship between `price` and `carat` of a diamond? What might explain that pattern? Run the code below and comment on the result.

```{r, eval=FALSE, echo=TRUE}
ggplot(diamonds) + 
  geom_point(aes(x = log(carat), y = log(price)), 
                color = "lightblue") +
  labs(title = "Price vs Carat")
```

7. What does the graph look like if we don't take logs? Is the relationship still linear? 

**Solution: When we remove the logs, the relationship is no longer linear. We see a large increase in price that is quite dispersed once the diamond is over 1 carat. We also notice bunching around round numbers of carats**


```{r, eval=TRUE, echo=TRUE}
ggplot(diamonds) + 
  geom_point(aes(x = carat, y = price), 
                color = "lightblue") +
  labs(title = "Price vs Carat")
```

8. We might want the specific statistical relationship. How much does log(price) increase on average when the log(carat) size increases by one unit? Enter statistics!

The function `lm()` runs a "linear regression" and answers our question: the log(price) increases by 1.676 log units for each log-carat.^[This is a preview of your stats sequence!] 


```{r, eval=FALSE}
lm(log(price) ~ log(carat), data = diamonds)
```

9. We can visualize this using the following code:

```{r, eval=FALSE, echo=TRUE}
ggplot(diamonds) + 
  geom_point(aes(x = log(carat), y = log(price)), 
                color = "lightblue") +
  # we get the estimates from linear regression output
  geom_abline(aes(slope = 1.676, intercept = 8.449)) + 
  labs(title = "Price vs Carat")
```

10. Is the relationship is stable across `cut`? Run a regression to find out how much does `log(price)` increase on average when the `log(carat)` size increases by one unit for __good cut diamonds__?  Make a plot showing the relationship.

```{r}
good_cut_diamonds <- diamonds[diamonds$cut == "Good",]

lm(log(price) ~ log(carat), data = good_cut_diamonds)

ggplot(good_cut_diamonds) + 
  geom_point(aes(x = log(carat), y = log(price)), 
                color = "lightblue") +
  geom_abline(aes(slope = 1.736, intercept =  8.376)) + 
  labs(title = "Price vs Carat: Good cut diamonds")
```




## We can make the above plots with base R as well.


```{r, eval=TRUE, echo=TRUE}
reg_model <- lm(log(price) ~ log(carat), data = diamonds)

plot(log(price) ~ log(carat), # plot log(price) by log(carat)
     type = "p", # plot type "p" (points)
     data = diamonds, # data from diamonds
     ylab = "log price", # add y-axis label
     xlab = "log carat", # add x-axis label
     main = "Price vs Carat" # add title
    )
abline(reg_model, col="green")
```


```{r, eval=TRUE, echo=TRUE}
plot(price ~ carat, # plot log(price) by log(carat)
     type = "p", # plot type "p" (points)
     data = diamonds, # data from diamonds
     ylab = "price", # add y-axis label
     xlab = "carat", # add x-axis label
     main = "Price vs Carat" # add title
    )
```

