---
title: 'TA Session 3: Intro to Normal Distribution'
author: "Harris Coding Camp -- Standard Track"
date: "Summer 2023"
output:
  beamer_presentation:
    fig_width: 6
    fig_height: 4
    includes:
      in_header: /Users/sheng-hao/Dropbox/uchicago/AIP 3rd Year/Coding Camp/Lecture
        Notes/Lecture 1/header_page_n.tex
---
```{r, echo=FALSE, include=FALSE}
library(tidyverse)
#library(readxl)
library(readr)
options(dplyr.summarise.inform = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, fig.align="center")
#knitr::opts_chunk$set(echo = TRUE, fig.align="center")
#housing_data <- read_csv("texas_housing_data.csv")
```

## Normal Distribution

  ![](/Users/sheng-hao/Dropbox/uchicago/AIP 2nd Year/Coding Camp/Lecture
        Notes/Lecture 6/images/6_3.png)

  - Most commonly used distribution; bell-curved shape
  - Important parameters: **mean & standard deviation**
    - Mean indicates where data points are centered
    - Standard deviation measures how data points are spread out

## Normal Distribution

  ![](/Users/sheng-hao/Dropbox/uchicago/AIP 2nd Year/Coding Camp/Lecture
        Notes/Lecture 6/images/6_4.png)

**Mean** is the average of a set of values. It also indicates where data points are centered.

## Normal Distribution

  ![](/Users/sheng-hao/Dropbox/uchicago/AIP 2nd Year/Coding Camp/Lecture
        Notes/Lecture 6/images/6_5.png)

**Variance** is a measure of how far a set of data are spread out from their mean value

  - **Standard deviation** is the sqrt of variance
  
## Normal Distribution and `rnorm()`
\footnotesize
At times, we feel the need to generate numbers at random to test certain hypothesis in statistics. Usually, we consider generating random numbers that follow a normal distribution using `rnorm()` function:

```{r, eval = FALSE}
rnorm(num, mean, sd)
```

  - `num`: number of values that needs to be generated
  - `mean`: mean value of the normal distribution. Its default value is 0.
  - `sd`: sd of the normal distribution. Its default value is 1.
  
```{r, eval = TRUE}
data <- rnorm(10)
data
```

## Normal Distribution and `rnorm()`
\footnotesize
Occationally, we consider generating random numbers that follow a normal distribution with specific mean and sd:
```{r, eval = TRUE}
data_1 <- rnorm(1000, 5, 2)
mean(data_1)
sd(data_1)
```

## Normal Distribution and `rnorm()`
\footnotesize
We can do simple simulation by generate 1000 random numbers using `rnorm()` and create a histogram based on that distribution:
```{r, eval = TRUE}
data <- rnorm(1000, 0, 1)
hist(data, main = "Normal distribution with mean = 0, sd = 1")
```

## Normal Distribution and `rnorm()`
\footnotesize
We can do simple simulation by generate 1000 random numbers using `rnorm()` and create a histogram based on that distribution:
```{r, eval = TRUE}
data <- rnorm(1000, 0, 10)
hist(data, main = "Normal distribution with mean = 0, sd = 10")
```