---
title: 'Accelerated Lecture 8: Iteration and Loops'
author: "Harris Coding Camp"
date: "Summer 2022"
output:
  beamer_presentation:
    includes:
      in_header: ../resources/header_page_n.tex
---
```{r, setup=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(readxl)
#options(dplyr.summarise.inform = FALSE)
#knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE, fig.align="center")
```

## Iteration and for-loops

We use for-loops to repeat a task over many different inputs or to repeat a simulation process several times.

- How to write for-loops
- When to use a for-loop vs vectorized code

```{r}
for(value in c(1, 2, 3, 4, 5)) {
  print(value^2)
}
```

## Simple for-loop

```{r}
for (x in c(3, 6, 9)) {
  print(x)
}
```

## Simple for-loop: what is going on?

```{r, eval = FALSE}
for (x in c(3, 6, 9)) {
  print(x)
}
```

The for-loop is equivalent to the following code. 

```{r, eval = FALSE}
x <- 3
print(x)
x <- 6
print(x)
x <- 9
print(x)
```

## Simple for-loop: what is going on?
\footnotesize
The variable (here `item`) is a **name** that you pick!

```{r, eval = FALSE}
for (item in c(3, 6, 9)) {
  print(item)
}
```

- it can be `anything`!!

```{r, eval = FALSE}
for (anything in c(3, 6, 9)) {
  print(anything)
}
```

## General structure of a `for` loop

The general structure of a `for` loop is as follows:

```{r eval=FALSE}
for (value in list_of_values) {
  do something (based on value)
}
```

Main components: Sequence, Body

## Components of a `for` loop

```{r eval=FALSE}
for (z in c(5, 4, 3)) { 
  print(z/2)
}
```

1. __Sequence__. Determines what to "loop over"
    - sequence above is `for (z in c(5, 4, 3))`
    - this creates a variable `z` 
    - we assign  `z` to values `c(5, 4, 3)` *iteratively*  
        - in the first iteration,  `z` is `5`
        - in the second iteration, `z` is `4`, etc.
2. __Body__. What to execute as we run through the loop.
    - Body in above loop is `print(z/2)`
    - Each iteration, the body prints the value of `z/2`

## The output.

for each value in `c(5,4,3)`, we divide it by 2.

```{r eval=TRUE}
for (z in c(5, 4, 3)) { 
  print(z/2)
}
```

*Of course, we have an easier way to divide items in a vector by 2*

```{r}
c(5, 4, 3) / 2
```

## Components of a `for` loop


1. __Sequence__. Determines what to "loop over"

  - often we loop over indices. 
  - recall, we can refer to items in a vector by their location
  
```{r eval=TRUE}
values <- c(5, 4, 3)

for (i in 1:3) { 
  print(values[[i]]/2)
}
```


## When to write a loop or use an iteration method

Grolemund and Wickham: __don't copy and paste more than twice__

- instead consider **a loop or function**

__Broadly, rationale for writing loop__:

- Can make changes to code in one place rather than many
- Easier to read


## When to write a loop vs a functions
\footnotesize
Loops are useful when:

  - a similar task is repeated many times in a row
  - you *cannot use a vectorized option*

Examples: 

    - read in data sets from individual years; each csv only differs by name

Functions are useful when:

  - we anticipate repeating tasks at different points in time
  - we require flexible and ad-hoc usage of the code
  - code is complex and naming and encapsulation helps clarify code functionality


Often we write functions and then put them in loops or other iterators!


## Recipe for how to write loop

The general recipe for writing a loop:

1. Complete the task for one instance outside a loop 

2. a) Decide which part(s) of the __body__ will change with each iteration

2. b) Write the __sequence__  

3. Usually you want to store the output, create an object to store the output outside of the loop

4. Construct the loop

## Example: find sample means
\footnotesize
Suppose we want to find the means of increasingly large samples.

```{r}
mean1 <- mean(rnorm(5))
mean2 <- mean(rnorm(10))
mean3 <- mean(rnorm(15))
mean4 <- mean(rnorm(20))
mean5 <- mean(rnorm(25000))

means <- c(mean1, mean2, mean3, mean4, mean5)
means
```



## Example: find sample means
\footnotesize
Let's avoid repeating code with a `for` loop. 

```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

for (i in seq_along(sample_sizes)) {
  sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
}

sample_means
```

In the following slides we'll explain each step.

## Finding sample means, broken down 

Assign initial variables **before** starting the for loop. 

```{r eval=FALSE}
# determine what to loop over
sample_sizes <- c(5, 10, 15, 20, 25000)

# pre-allocate space to store output
sample_means <- rep(0, length(sample_sizes))
```

## why pre-allocate space?

\footnotesize

```{r, eval=FALSE}
# pre-allocate space to store output
sample_means <- rep(0, length(sample_sizes))

# inside for-loop
# same data object every step
sample_means[[i]] <- next_item
```


 - The alternative is to build up an object as you go (can be very slow)
 
```{r, eval=FALSE}
# before
sample_means <- c()

# inside for-loop 
# new data object every step
sample_means <- c(sample_means, next_item)
```




## Reviewing alternative ways to pre-allocate space

Each data type has a comparable function e.g. `logical()`, `integer()`, `character()`.


```{r, eval=FALSE}
# All equivalent
sample_means <- rep(0.0, 5)
sample_means <- vector("double", length = 5)
sample_means <- numeric(5)
sample_means <- double(5)
```


To hold data of different types, we'll use lists.

```{r, eval=FALSE}
data_list <- vector("list", length = 5)
```

## Adding data to a vector, broken down 

Determine what sequence to loop over. 

- we iterate over the indices! 
- Numbers from 1 to length(sample_sizes)

```{r eval=FALSE}
for (i in 1:length(sample_sizes)) {

}
```

## A helper function `seq_along()`

`seq_along(x)` is synonymous to `1:length(x)` 

where `x` is a vector.

**Simple Example**
```{r}
vec <- c("x", "y", "z")
1:length(vec)
seq_along(vec)
```

## A helper function `seq_along()`

`seq_along()` protects against that moment when `length(x) = 0`

  - you might worry about this in a function when you don't have control over the input.

```{r}
seq_along(NULL)
```

```{r}
# equivalent to 1:0
1:length(NULL) 
```


## A helper function `seq_along()`

**Back to Our Example**
```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
1:length(sample_sizes)
seq_along(sample_sizes)
```

## Adding data to a vector, broken down 
\footnotesize
Add `for`-loop structure:
```{r eval=FALSE}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

for (i in seq_along(sample_sizes)) {

}
```

When writing loops, it's very common to create a **sequence** from 1 to the length (i.e., number of elements) of an object. 

- __sequence__: `for (i in seq_along(sample_sizes))`
  - `i` takes on 1, 2, 3, 4 and 5 *sequentially*
  - Sequence iterates through the _position number_ or _index_ of each element in `sample_sizes`

## Adding data to a vector, broken down
\footnotesize
Add `for`-loop body:
```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

for (i in seq_along(sample_sizes)) {
  sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
}
```

__body__: value of `i` refers to the _position number_ or _index_ of the $i^{th}$ element in `sample_sizes`

- Access element contents using `sample_sizes[[i]]`
- Here, save the output as the $i^{th}$ element in `sample_means`

## Adding data to a vector, broken down
\footnotesize
Now `sample_means` has stored the results of iteratively running our code!

```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

for (i in seq_along(sample_sizes)) {
  sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
}

sample_means
```

## To belabor the point
\footnotesize
Our code is equivalent to ...

```{r}
sample_sizes <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(sample_sizes))

i <- 1
sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
i <- 2
sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
i <- 3
sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
i <- 4
sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))
i <- 5
sample_means[[i]] <- mean(rnorm(sample_sizes[[i]]))


sample_means
```


## You try: Why doesn't this work?

- Run the code
- Try to fix it.

```{r, eval = FALSE}
n <- 1:5
sample_means <- rep(0, length(n))

for (i in seq_along(n)) {
  mean(rnorm(n[[i]]))
}

sample_means
```

## Aside: Common errors
\footnotesize
This code falls, why?

- It's not *not* running!

```{r}
n <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(n))

for (i in seq_along(n)) {
   print(mean(rnorm(n[[i]])))
}

sample_means
```



## Aside: Common errors
\footnotesize
This code falls, why?

- It's running!
- But we didn't save the output!

```{r}
n <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(n))

for (i in seq_along(n)) {
  sample_means[[i]] <- mean(rnorm(n[[i]]))
}

sample_means
```

## Aside: Common errors
\footnotesize
What's wrong with this code?

```{r}
n <- c(5, 10, 15, 20, 25000)
sample_means <- rep(0, length(n))

for (i in seq_along(n)) {
 sample_means[[i]] <-  mean(rnorm(n[[1]]))
}

sample_means
```



## Aside: Debugging with `cat()`
\footnotesize
`cat()` also prints out the output - and can handle variables!

```{r}
for (i in seq_along(n)) {
  
  sample_means[[i]] <- mean(rnorm(n[[i]]))
  
  cat("mean of sample", i, "is", sample_means[[i]], 
      fill = TRUE)
}
```

## Aside: Debugging with `cat()`
\footnotesize
What does `fill=TRUE` do? 
```{r}
for (i in seq_along(n)) {
  
  sample_means[[i]] <- mean(rnorm(n[[i]]))
  cat("mean of sample", i, "is", sample_means[[i]])
}
```

## Aside: Debugging with `cat()`
\footnotesize
We can explicitly make a new line with "`\n`" or use `fill = TRUE`.

```{r}
for (i in seq_along(n)) {
  
  sample_means[[i]] <- mean(rnorm(n[[i]]))
  cat("mean of sample", i, "is", sample_means[[i]], "\n")
}
```
## Aside: Debugging with `cat()`

`cat()` prints vectors directly

- i.e. it's not vectorized

```{r}
names <- c("Sahil", "Kate", "Andrew")
cat("Our TA", names, "is great!", fill=TRUE)

# paste separates input with " " by default
# paste0 does no separation.
print(paste("Our TA", names, "is great!"))
```



```{r, eval = FALSE}
# What about print by itself? 
# ... throws an ERROR
print("Our TA", names, "is great!")
```

## Try it yourself

1. Create a numeric vector that has year of birth of members of your family
    - you decide who to include
    - e.g., `birth_years <- c(1944, 1950, 1981, 2016)`
2. Write a loop that calculates the age of each member of you family.
3. Print the output in sentences
    - e.g. The age of family member `i` is...
    
*Note: multiple correct ways to complete this task*
    
4. Write the same code vectorized.

\medskip


```{r, echo=FALSE, eval=FALSE}
# answer
birth_years <- c(1944,1950,1981,2016)
family_members <- numeric(length(birth_years))

for(i in seq_along(family_members)) { # Loop sequence
  age <- 2022 - birth_years[[i]]        # Loop body
  cat("The age of family member",i,"is",age, fill=TRUE)
}
```

## Review: Vectorized operations

When possible, take advantage of vectorization!

```{r}
a <- 7:11
b <- 8:12
out <- rep(0L, 5)

for (i in seq_along(a)) {
  out[[i]] <- a[[i]] + b[[i]]
}

out
```

This is a bad example of a for loop!

## The better alternative: vectorized addition

```{r}
a <- 7:11
b <- 8:12
out <- a + b

out
```

Use vectorized operations when you can.

- easier to read code
- easier to write code (eventually!)



## What happens when we loop over a tibble?
\footnotesize

```{r}
df <- tibble(a = rnorm(4), b = rnorm(4))
df
```

```{r}
for (i in seq_along(df)) {
  cat("value of object", i, "=", df[[i]], "\n") 
}
```

We loop over columns, *not* rows!

- recall tibbles are `list`s and each entry in the list is a column!


## It unnatural to loop over *rows* 

We have vectorized functions with mutate  or `$<-`

```{r, eval = FALSE}
data %>%
  mutate(new_col = something_vectorized(old_col))

data$new_col <- something_vectorized(old_col1, old_col2)
```

and `rowwise()` for unvectorized code

```{r, eval = FALSE}
data %>%
  rowwise() %>%
  mutate(new_col = something_unvectorized(old_col))
```

And, of course, you can pull out a column as a vector and iterate over it.


## An example of iterating over columns

Task: calculates z-scores for a set of variables in a data frame

\footnotesize
First, create sample data
```{r, warning = FALSE}
# matrix is like a 2d atomic vector
set.seed(4)
df <- as_tibble(matrix(runif(40), ncol = 4))
names(df) <- c("a", "b", "c", "d")
head(df)
```



## An Example of iterating over columns
\footnotesize
The z-score for observation _i_ is the number of standard deviations from mean:

$z_i = \frac{x_i - \bar{x}}{sd(x)}$

Let's calculate z-score for first 4 observations of `df$a`:
```{r}
(df$a[1] - mean(df$a, na.rm=TRUE))/sd(df$a, na.rm=TRUE)
(df$a[2] - mean(df$a, na.rm=TRUE))/sd(df$a, na.rm=TRUE)
(df$a[3] - mean(df$a, na.rm=TRUE))/sd(df$a, na.rm=TRUE)
```

## Hmm ... Maybe we need a function!

```{r}
calc_z_score <- function(x, i, na.rm = TRUE) {
  (x[i] - mean(x, na.rm = na.rm)) / sd(x, na.rm = na.rm)
}

calc_z_score(df$a, 1)
calc_z_score(df$a, 2)
calc_z_score(df$a, 3)
```

## Hmm ... Maybe we can get rid of `i` with vectorized code

```{r}
calc_z_score <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

calc_z_score(df$a)
```



## Example of modifying an object: z-score loop
\footnotesize
__Our Task__: write loop that replaces variables with z-scores of those variables

\medskip

- __sequence__ 
    - data frame `df` has 4 variables and all are quantitative
    - operate on each column
        - `for (i in seq_along(df))`
- __body__
    - Take z-score function:
        - `calc_z_score(x)`
    - Replace `x` with `df[[i]]`: 
        - `calc_z_score(df[[i]])`
    - Assign or replace each column: 
        - overwrite: `df[[i]] <- calc_z_score(df[[i]])`
        - make new object: `out_df[[i]] <- calc_z_score(df[[i]])`

## Example of modifying an object: z-score loop

Creating an object to capture output is a bit more involved with data frames.
\footnotesize
```{r}
# we can use the old object
out_df <- df

# OR we can make an empty dataframe
out_df <- as_tibble(matrix(rep(NA, 40), ncol = ncol(df)))
names(out_df) <- names(df)

head(out_df)
```

##  The whole loop

```{r}
out_df <- as_tibble(matrix(rep(0, 40), ncol = ncol(df)))
names(out_df) <- names(df)

for (i in seq_along(df)) {
  # modify values
  out_df[[i]] <- calc_z_score(df[[i]])
}
str(out_df)
```


## Modifying an object in place

We can also change df in place!

- Useful if df is very large (relative to your RAM)
- Theoretically, can do this in other loops we've seen, but dangerous!
  - We might change underlying data that we operate on in the next iteration!

\footnotesize
```{r}
for (i in seq_along(df)) {
  # modify values
  df[[i]] <- calc_z_score(df[[i]])
}
str(df)
```

## map or apply

 Many R coders prefer the `map()` family functions from `purrr` or base R `apply` family. 
 
 - See [iteration](https://r4ds.had.co.nz/iteration.html) in R for Data Science 

```{r, eval = FALSE}
# map(.x, .f)
map(df, calc_z_score)
# sapply(X, FUN, ..., simplify = TRUE)
sapply(df, calc_z_score, simplify = FALSE)
```

This says "apply" the function to the columns of the `df` or "map" the columns of `df` to the function `calc_z_score`.

Output is a list -- here, a list of modified columns.

## In action
\footnotesize
```{r}
# map_<output type>(.x, .f)
map(df, calc_z_score) %>% bind_cols() %>% head(4)

sapply(df, calc_z_score, simplify = FALSE) %>% 
  bind_cols() %>% head(4)
```

## `map` makes a list, but we usually want vectors or tibbles.

The `map` family has the form 

- `map_<output type>(.x, .f)`

```{r, eval = FALSE}
map(.x, .f) %>% as.integer()
map_int(.x, .f)

map(.x, .f) %>% as.character()
map_chr(.x, .f)
```
```{r}
# dfc = data.frame columns
# map(df, calc_z_score) %>% bind_cols()
map_dfc(df, calc_z_score) %>% head()
```

## `map` functions can feel like magic

- imagine writing a loop 

  - the sequence is `.x`
  - the body is `.f` 
    - often you'll write new functions or even use "anonymous functions".
    
```{r}
sample_means <- map_dbl(c(1, 10, 100, 1000),
                        function(x) mean(rnorm(x)))
sample_means
```

- There's much less overhead. 


## `sapply` can feel like magic

- imagine writing a loop 

  - the sequence is `X`
  - the body is `FUN` 
    - often you'll write new functions or even use "anonymous functions".
    
```{r}
sample_means <- sapply(c(1, 10, 100, 1000),
                        function(x) mean(rnorm(x)))
sample_means
```

- There's much less overhead. 
- By default  `simplify = TRUE`, so `sapply` outputs a double vector. 
  - This "simplification" can make for confusing code since you might be surprised by the output. 

## Key points: iteration

- Iteration is useful when we are repeatedly calling the same block of code or function while changing one (or two) inputs. 

- If you can, use vectorized operations.

- Otherwise, for loops work for iteration

    - Clearly define what you will iterate over (values or indicies)
    - Pre-allocate space for your output
    - The body of the for-loop has parametrized code based on thing your iterating over
    - Debug as you code by testing your understanding of what the for-loop should be doing (e.g. using `cat()` or `print()`)
    

## Next steps

Lab:

- *Today:* Learning Loops

**I can write loops, but know when to vectorize**

- *Tomorrow:* Joining data sets, Base R plots, Review / Final Project
- *Friday:*  Pivoting, Review / Final Project

Final project:

- Deadline for guaranteed feedback September 24. 
  - Recommend setting a personal deadline of the 21st.
  - Optional but worth trying!  
  
# Thank you!




# Additional Material

## Creating multiple plots with a loop
\footnotesize
Another good use of a loop is to create multiple graphs easily. Let’s use a loop to create 4 plots representing data from an exam containing 4 questions. Here are how the first few rows of the data look:

```{r echo=FALSE}
examscores <- tibble(a = rnorm(100,60,20), 
                     b = rnorm(100,50,25), 
                     c = rnorm(100,70,15),
                     d = rnorm(100,75,10))
```

```{r}
head(examscores)
```

## Creating multiple plots with a loop
\footnotesize

Let's loop over the columns and create a histogram of the data in each column:
```{r, eval=FALSE}
# Set up a 2 x 2 plotting space
par(mfrow = c(2, 2))  

# Create the loop.vector (sequence)
each.question <- 1:4

for (i in each.question) { 
  
  # Plot histogram of each question
  hist(examscores[[i]],
       main = paste("Question", i),
       xlab = "Scores",
       xlim = c(0, 100))
}
```

## Creating multiple plots with a loop

```{r, echo=FALSE}
# Set up a 2 x 2 plotting space
par(mfrow = c(2, 2))  

# Create the loop.vector (sequence)
each.question <- 1:4

for (i in each.question) { 
  
  # Plot histogram of each question
  hist(examscores[[i]],
       main = paste("Question", i),
       xlab = "Scores",
       xlim = c(0, 100))
}
```

## Try it yourself
\footnotesize
We’ll use `midwest` data and use a loop to create 4 plots representing data from `midwest` containing 4 key columns: `poptotal`, `percpovertyknown`, `percollege` and `percbelowpoverty`. Loop over these selected columns and create a histogram of the data in each column by completing the following code.

```{r, eval=FALSE}
# Set up a 2 x 2 plotting space
par(mfrow = c(2, 2))  

# Create the sequence
selected.column <- c("poptotal", "percpovertyknown", 
                     "percollege", "percbelowpoverty")

for (...) { 
  
  # Plot histogram of each column
  hist(...,
       main = paste(...),
       xlab = ...)
}
```

## Try it yourself

You should get the histograms below:

\medskip 

```{r, echo=FALSE}
# answer
# Set up a 2 x 2 plotting space
par(mfrow = c(2, 2))  

# Create the sequence
selected.column <- c("poptotal", "percpovertyknown", "percollege", "percbelowpoverty")

for (i in selected.column) { 
  
  # Plot histogram of each question
  hist(midwest[[i]],
       main = paste("Histogram of", i),
       xlab = i)
}
```
